---
title: "ST 558 Project 1"
author: "Grace Haley"
format: html
editor: visual
---
## Introduction

For this project, I am using data from the Census PUMS API. I will be creating functions to pull and clean the data, summarize different variables, and make plots. I also want the functions to work with different years and geographic areas so I can easily compare the data in different ways.

## Part 1: Data Processing

### Parsing Variable Values

Before pulling the actual PUMS data, I first need some information about the variables I will be using. The Census API has JSON files that give the possible values and descriptions for each variable. I will save this information so I can use it later when cleaning the data.

```{r}
library(httr)
library(jsonlite)
library(tidyverse)
library(lubridate)
```

I am starting with 2024 so I can see how the variable information is set up before making a function that will work for all four years.

```{r}
url_2024 <- "https://api.census.gov/data/2024/acs/acs1/pums/variables.json"

response_2024 <- GET(url_2024)

# Parse the JSON response into an R list
variables_2024 <- content(response_2024, as = "parsed")

# Variables needed for 2024
needed_vars <- c(
  "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
  "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
  "STATE", "REGION", "DIVISION"
)

# Keep only the variables needed for the project
selected_2024 <- variables_2024$variables[needed_vars]

names(selected_2024)
```

Now that I know how the variable information is stored, I can make a function to do this for each year. One thing I have to account for is that the state variable is called ST in 2021 and 2022, but STATE in 2023 and 2024.

```{r}
# Pull and parse the variable information for a given year
get_variable_info <- function(year) {
  
  # Create the URL for the selected year
  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums/variables.json"
  )
  
  # Pull and parse the JSON data
  response <- GET(url)
  variable_data <- content(response, as = "parsed")
  
  # The state variable has a different name in 2021 and 2022
  state_var <- if (year %in% c(2021, 2022)) "ST" else "STATE"
  
  # Variables needed for the project
  needed_vars <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
    "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
    state_var, "REGION", "DIVISION"
  )
  
  # Return only the needed variables
  variable_data$variables[needed_vars]
}
```
The function works for both versions of the state variable, so I can use it 
for all four years. I will save everything together in the data folder so I 
can use it again later without having to pull it every time.

```{r}
# Get the variable information for each year
variable_info_2021 <- get_variable_info(2021)
variable_info_2022 <- get_variable_info(2022)
variable_info_2023 <- get_variable_info(2023)
variable_info_2024 <- get_variable_info(2024)

# Store all four years together in one list
variable_info <- list(
  `2021` = variable_info_2021,
  `2022` = variable_info_2022,
  `2023` = variable_info_2023,
  `2024` = variable_info_2024
)

# Save the variable information in the data folder
saveRDS(variable_info, "data/variable_info.rds")
```

### Parsing API Data

Now I can start working with the actual PUMS data. I am going to start with a simple API request to see what the returned data looks like and then create a function to turn it into a tibble.

```{r}
# Test a basic PUMS API call
pums_response <- GET(
  "https://api.census.gov/data/2024/acs/acs1/pums",
  query = list(
    get = "AGEP,PWGTP,SEX",
    `for` = "state:37",
    key = Sys.getenv("CENSUS_API_KEY")
  )
)

# Check that the request was successful
status_code(pums_response)
```

The API returns the variable names in the first row and the actual data in the rows after that. I will make a small helper function to separate those and return the data as a tibble.

```{r}
# Convert a Census API response into a tibble
response_to_tibble <- function(response) {
  
  # Read the JSON returned by the API
  api_data <- content(
    response,
    as = "parsed",
    type = "application/json",
    encoding = "UTF-8",
    simplifyVector = FALSE
  )
  
  # The first row has the column names
  column_names <- unlist(api_data[[1]])
  
  # Everything after the first row is the actual data
  data_rows <- lapply(api_data[-1], unlist)
  data_matrix <- do.call(rbind, data_rows)
  
  # Turn everything into a tibble and add the column names
  census_tibble <- as_tibble(
    data_matrix,
    .name_repair = "minimal"
  )
  
  names(census_tibble) <- column_names
  
  census_tibble
}
```

The helper function gets the data into a tibble, but the variables still need to be converted to the right types. I will handle that in the main function.

JWAP and JWDP are a little different from the other numeric variables because the Census stores them as codes for ranges of time. I will use the middle of each time range as the value, so I made a helper function to calculate the midpoint.

```{r}
# Convert a Census time range into its midpoint
get_time_midpoint <- function(time_range) {
  
  # Return NA if there is no time to use
  if (
    length(time_range) == 0 ||
    is.na(time_range) ||
    time_range == ""
  ) {
    return(NA)
  }
  
  # Clean up a.m. and p.m. so the times can be read
  time_range <- str_replace_all(
    time_range,
    c("a\\.m\\." = "AM", "p\\.m\\." = "PM")
  )
  
  # Split the beginning and ending times
  times <- str_split(time_range, " to ")[[1]]
  
  # Read both times using lubridate
  start_time <- parse_date_time(times[1], orders = "I:M p")
  end_time <- parse_date_time(times[2], orders = "I:M p")
  
  # Find the middle of the time range
  midpoint <- start_time + (end_time - start_time) / 2
  
  midpoint
}
```

Next, I will make the main function for pulling the PUMS data. I want the function to let me choose the year, variables, and geographic area while also checking that the inputs are valid. It will also take care of converting the variables to the right types.

```{r}
get_pums_data <- function(
  year = 2024,
  numeric_vars = c("AGEP", "PWGTP"),
  categorical_vars = "SEX",
  geography = "State",
  geography_value = NULL
) {
  
  # Make sure the year is one of the four years being used
  valid_years <- c(2021, 2022, 2023, 2024)
  
  if (!year %in% valid_years) {
    stop("Year must be 2021, 2022, 2023, or 2024.")
  }
  
  # Check the numeric variables
  valid_numeric_vars <- c(
    "AGEP", "GASP", "GRPIP", "JWAP",
    "JWDP", "JWMNP", "PWGTP"
  )
  
  if (!all(numeric_vars %in% valid_numeric_vars)) {
    stop("One or more numeric variables are invalid.")
  }
  
  if (!any(numeric_vars != "PWGTP")) {
    stop("At least one numeric variable besides PWGTP must be selected.")
  }
  
  # PWGTP always needs to be included
  numeric_vars <- unique(c(numeric_vars, "PWGTP"))
  
  # Check the categorical variables
  valid_categorical_vars <- c(
    "FER", "HHL", "SCH", "SCHL", "SEX"
  )
  
  if (length(categorical_vars) == 0) {
    stop("At least one categorical variable must be selected.")
  }
  
  if (!all(categorical_vars %in% valid_categorical_vars)) {
    stop("One or more categorical variables are invalid.")
  }
  
  # Make sure only one geography level was chosen
  valid_geographies <- c("Region", "Division", "State")
  
  if (length(geography) != 1) {
    stop("Exactly one geography level must be selected.")
  }
  
  if (!geography %in% valid_geographies) {
    stop("Geography must be Region, Division, or State.")
  }
  
  # Load the variable information saved earlier
  variable_info <- readRDS("data/variable_info.rds")
  year_info <- variable_info[[as.character(year)]]
  
  # Set up the selected geography
  if (geography == "State") {
    geography_var <- if (year %in% c(2021, 2022)) "ST" else "STATE"
    geography_values <- year_info[[geography_var]]$values$item
    default_value <- "37"
    
  } else if (geography == "Region") {
    geography_var <- "REGION"
    geography_values <- year_info$REGION$values$item
    default_value <- "3"
    
  } else {
    geography_var <- "DIVISION"
    geography_values <- year_info$DIVISION$values$item
    default_value <- "5"
  }
  
  # Use the default if a specific geography was not given
  if (is.null(geography_value)) {
    geography_value <- default_value
  }
  
  # Make sure the geography value is actually available
  if (
    geography_value != "all" &&
    !geography_value %in% names(geography_values)
  ) {
    stop("The selected geography value is invalid.")
  }
  
  # Put together the variables we want from the API
  requested_vars <- unique(c(
    numeric_vars,
    categorical_vars,
    geography_var
  ))
  
  # Build the API URL for the selected year
  api_url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums"
  )
  
  # Census uses * when we want every value for a geography
  api_geography_value <- if (
    geography_value == "all"
  ) {
    "*"
  } else {
    geography_value
  }
  
  geography_query <- paste0(
    tolower(geography),
    ":",
    api_geography_value
  )
  
  # Pull the data from the Census API
  response <- GET(
    api_url,
    query = list(
      get = paste(requested_vars, collapse = ","),
      `for` = geography_query,
      key = Sys.getenv("CENSUS_API_KEY")
    )
  )
  
  # Turn the API response into a tibble
  census_data <- response_to_tibble(response)
  
  # Convert the regular numeric variables
  regular_numeric_vars <- setdiff(
    numeric_vars,
    c("JWAP", "JWDP")
  )
  
  census_data <- census_data |>
    mutate(
      across(
        all_of(regular_numeric_vars),
        as.numeric
      )
    )
  
  # See if either of the time variables was selected
  time_vars <- intersect(
    numeric_vars,
    c("JWAP", "JWDP")
  )
  
  # Convert the time codes to the middle of each time range
  for (var in time_vars) {
    
    time_values <- year_info[[var]]$values$item
    
    time_ranges <- sapply(
      census_data[[var]],
      function(code) {
        
        # A code of 0 means there is no time to use
        if (is.na(code) || code == "" || code == "0") {
          return(NA_character_)
        }
        
        # Add leading zeros so the codes match the saved information
        code <- str_pad(
          code,
          width = 3,
          side = "left",
          pad = "0"
        )
        
        # If there still is not a match, treat it as missing
        if (!code %in% names(time_values)) {
          return(NA_character_)
        }
        
        unlist(time_values[[code]])
      }
    )
    
    census_data[[var]] <- do.call(
      c,
      lapply(
        time_ranges,
        get_time_midpoint
      )
    )
  }
  
  # Replace the category codes with their actual labels
  for (var in categorical_vars) {
    
    category_values <- year_info[[var]]$values$item
    
    census_data[[var]] <- factor(
      census_data[[var]],
      levels = names(category_values),
      labels = unname(unlist(category_values))
    )
  }
  
  # The API adds the geography again, so remove that extra column
  api_geography_var <- tolower(geography)
  
  if (
    api_geography_var %in% names(census_data) &&
    api_geography_var != geography_var
  ) {
    census_data <- census_data |>
      select(-all_of(api_geography_var))
  }
  
  # Give the finished tibble the census class
  class(census_data) <- c(
    "census",
    class(census_data)
  )
  
  census_data
}
```
### Combining Years

The function above only pulls one year at a time, but I also want to be able to look at multiple years together. I will make another function that uses the single-year function for each year I choose and then combines the results. I will also add a year column so I can tell which year each observation came from.

```{r}
# Pull PUMS data for more than one year
get_multiple_years <- function(
  years = "all",
  numeric_vars = c("AGEP", "PWGTP"),
  categorical_vars = "SEX",
  geography = "State",
  geography_value = NULL
) {
  
  # These are the years available for this project
  valid_years <- c(2021, 2022, 2023, 2024)
  
  # Use all four years if "all" is selected
  if (length(years) == 1 && years == "all") {
    years <- valid_years
  }
  
  # Make sure all of the selected years are available
  if (!all(years %in% valid_years)) {
    stop("Years must be 2021, 2022, 2023, 2024, or all.")
  }
  
  # Pull the data for each selected year
  year_data <- lapply(
    years,
    function(current_year) {
      
      data <- get_pums_data(
        year = current_year,
        numeric_vars = numeric_vars,
        categorical_vars = categorical_vars,
        geography = geography,
        geography_value = geography_value
      )
      
      # Keep the state variable name the same across years
      if (geography == "State" && "ST" %in% names(data)) {
        data <- data |>
          rename(STATE = ST)
      }
      
      # Keep track of which year each row came from
      data$year <- current_year
      
      data
    }
  )
  
  # Combine all of the years into one tibble
  combined_data <- bind_rows(year_data)
  
  # Treat year as a category instead of a numeric variable
  combined_data$year <- factor(combined_data$year)
  
  # Keep the census class after combining the data
  class(combined_data) <- c(
    "census",
    setdiff(class(combined_data), "census")
  )
  
  combined_data
}
```

## Part 2: Generic Functions

### Summary Function

Now that the data can be pulled and combined, I want an easier way to summarize it. I will create a summary function specifically for census objects. For numeric variables, I will use the person weights to calculate weighted means and standard deviations. For categorical variables, I will use the weights to get counts for each category.

```{r}
# Summarize numeric and categorical Census variables
summary.census <- function(
  object,
  numeric_vars = NULL,
  categorical_vars = NULL,
  ...
) {
  
  # Don't include the weights or time variables in the numeric summaries
  ignored_numeric <- c("PWGTP", "JWAP", "JWDP")
  
  # Find the numeric variables if none were selected
  if (is.null(numeric_vars)) {
    numeric_vars <- names(object)[
      sapply(object, is.numeric)
    ]
    
    numeric_vars <- setdiff(
      numeric_vars,
      ignored_numeric
    )
  }
  
  # Find the categorical variables if none were selected
  if (is.null(categorical_vars)) {
    categorical_vars <- names(object)[
      sapply(object, is.factor)
    ]
  }
  
  # This list will hold all of the summaries
  summary_results <- list()
  
  # Calculate weighted means and standard deviations
  for (var in numeric_vars) {
    
    values <- object[[var]]
    weights <- object$PWGTP
    
    keep <- !is.na(values) & !is.na(weights)
    values <- values[keep]
    weights <- weights[keep]
    
    weighted_mean <- sum(values * weights) / sum(weights)
    
    weighted_sd <- sqrt(
      sum(values^2 * weights) / sum(weights) -
        weighted_mean^2
    )
    
    summary_results[[var]] <- c(
      mean = weighted_mean,
      sd = weighted_sd
    )
  }
  
  # Calculate weighted counts for categorical variables
  for (var in categorical_vars) {
    
    values <- object[[var]]
    weights <- object$PWGTP
    
    weighted_counts <- tapply(
      weights,
      values,
      sum,
      na.rm = TRUE
    )
    
    summary_results[[var]] <- weighted_counts
  }
  
  # Return all of the summaries
  summary_results
}
```

### Plot Function

I also want a quick way to compare a numeric variable across different categories. I will create a plot function for census objects that makes a weighted boxplot using the person weights.

```{r}
# Make a weighted boxplot for a numeric and categorical variable
plot.census <- function(
  x,
  categorical_var,
  numeric_var,
  ...
) {
  
  # Make sure the selected categorical variable is a factor
  if (!is.factor(x[[categorical_var]])) {
    stop("The categorical variable must be a factor.")
  }
  
  # Make sure the selected numeric variable is numeric
  if (!is.numeric(x[[numeric_var]])) {
    stop("The numeric variable must be numeric.")
  }
  
  # Create the weighted boxplot
  ggplot(
    x,
    aes(
      x = get(categorical_var),
      y = get(numeric_var),
      weight = PWGTP
    )
  ) +
    geom_boxplot() +
    labs(
      x = categorical_var,
      y = numeric_var
    )
}
```

## Investigation

Now that all of the functions are working, I can use them to look at the PUMS data in a few different ways. I am going to compare some of the variables across years and groups and use the summary and plot functions I created above to look for any differences or patterns in the data.

### Commuting Patterns in North Carolina

I decided to look at commuting patterns in North Carolina and see if there are any differences across years or between males and females. I will start by pulling age and travel time to work for all four years.

```{r}
# Pull North Carolina commuting data for all four years
commute_data <- get_multiple_years(
  years = "all",
  numeric_vars = c("AGEP", "JWMNP", "PWGTP"),
  categorical_vars = "SEX",
  geography = "State",
  geography_value = "37"
)

glimpse(commute_data)
```

First, I will use my summary function to get an overall look at the people with a reported travel time and see how the data are split across sex and year. Since JWMNP has a lot of values equal to 0, I am going to leave those out when looking specifically at commute times. This way the commute time summaries are based on people who have a reported travel time greater than 0.

```{r}
# Keep people with a reported travel time
commute_workers <- commute_data |>
  filter(JWMNP > 0)

# Keep the census class after filtering
class(commute_workers) <- c(
  "census",
  class(commute_workers)
)

commute_summary <- summary(commute_workers)

commute_summary
```

After removing the 0 travel times, the average commute time is about 25.2 minutes. The standard deviation is about 21.1 minutes, so there is quite a bit of variation in how long people are commuting. The average age for this group is about 41.6 years. The weighted counts also increase across the four years, with the largest count in 2024.

I also want to see if commute times look different between males and females. I can use my plot function to compare the distributions.

```{r}
# Compare commute times by sex
plot(
  commute_workers,
  categorical_var = "SEX",
  numeric_var = "JWMNP"
)
```

The commute time distributions for males and females look pretty similar. Their medians are almost the same and most of the values fall within a similar range. Both groups also have quite a few longer commute times that show up as outliers, so there does not seem to be a major difference between the two groups from this plot.

Next, I want to see if commute times changed much across the four years.

```{r}
# Compare commute times across years
plot(
  commute_workers,
  categorical_var = "year",
  numeric_var = "JWMNP"
)
```

The commute time distributions stay pretty similar from 2021 through 2024. The medians are close across all four years, although 2024 looks slightly higher. There are also longer commute time outliers in every year, so overall the distributions do not seem to have changed very much.

To get a more exact comparison, I will also use my summary function separately for each year.

```{r}
# Get the commute time summary for each year
year_summaries <- lapply(
  levels(commute_workers$year),
  function(current_year) {
    
    year_data <- commute_workers |>
      filter(year == current_year)
    
    class(year_data) <- c(
      "census",
      class(year_data)
    )
    
    summary(
      year_data,
      numeric_vars = "JWMNP",
      categorical_vars = character(0)
    )
  }
)

names(year_summaries) <- levels(commute_workers$year)

year_summaries
```

The weighted average commute time increased a little each year. It went from 
about 24.6 minutes in 2021 to 25.8 minutes in 2024, which is an increase of 
a little over one minute. This also matches the boxplots above, where the 
commute times looked pretty similar across the four years but were slightly 
higher in 2024.

### Age and School Enrollment in the South

For one more comparison, I wanted to try the functions with a different 
geographic level and different variables. This time, I will look at age and 
school enrollment in the South using the 2024 data.

```{r}
# Pull 2024 data for the South
south_data <- get_pums_data(
  year = 2024,
  numeric_vars = c("AGEP", "PWGTP"),
  categorical_vars = "SCH",
  geography = "Region",
  geography_value = "3"
)

summary(south_data)
```

The weighted average age in the South in 2024 was about 39.5 years. The 
largest school enrollment group was people who had not attended school in 
the last three months. There were also more people attending public school 
or public college than private school, college, or home school.

### Investigation Takeaways

Overall, the functions made it pretty easy to look at the Census data in a 
few different ways without having to rewrite the same code each time. For 
the North Carolina commuting data, average travel time to work increased 
slightly from 2021 to 2024, although the distributions were pretty similar 
across the four years. The comparison by sex also showed similar commute 
time distributions for males and females.

I also tried the functions with a different geographic level by looking at 
the South in 2024. The weighted average age was about 39.5 years, and the 
largest school enrollment group was people who had not attended school in 
the last three months. This was a useful way to check that the functions 
could handle different variables and geographic selections instead of only 
the North Carolina example.