---
title: "ST 558 Project 1"
author: "Grace Haley"
format: html
editor: visual
---
## Introduction

For this project, I am using data from the Census PUMS API. I will be creating functions to pull and clean the data, summarize different variables, and make plots. I also want the functions to work with different years and geographic areas so I can easily compare the data in different ways.

## Part 1: Data Processing

### Parsing Variable Values

Before pulling the actual PUMS data, I first need some information about the variables I will be using. The Census API has JSON files that give the possible values and descriptions for each variable. I will save this information so I can use it later when cleaning the data.

```{r}
library(httr)
library(jsonlite)
library(tidyverse)
library(lubridate)
```

I am starting with 2024 so I can see how the variable information is set up before making a function that will work for all four years.

```{r}
url_2024 <- "https://api.census.gov/data/2024/acs/acs1/pums/variables.json"

response_2024 <- GET(url_2024)

response_2024

# Parse the JSON response into an R list
variables_2024 <- content(response_2024, as = "parsed")

# Look at the names of the first few variables
head(names(variables_2024$variables))

# Look at the information stored for the SEX variable
variables_2024$variables$SEX

# Pull out the possible values for SEX
variables_2024$variables$SEX$values$item

# Variables needed for 2024
needed_vars <- c(
  "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
  "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
  "STATE", "REGION", "DIVISION"
)

# Keep only the variables needed for the project
selected_2024 <- variables_2024$variables[needed_vars]

names(selected_2024)
```

Now that I know how the variable information is stored, I can make a function to do this for each year. One thing I have to account for is that the state variable is called ST in 2021 and 2022, but STATE in 2023 and 2024.

```{r}
# Pull and parse the variable information for a given year
get_variable_info <- function(year) {
  
  # Create the URL for the selected year
  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums/variables.json"
  )
  
  # Pull and parse the JSON data
  response <- GET(url)
  variable_data <- content(response, as = "parsed")
  
  # The state variable has a different name in 2021 and 2022
  state_var <- if (year %in% c(2021, 2022)) "ST" else "STATE"
  
  # Variables needed for the project
  needed_vars <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
    "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
    state_var, "REGION", "DIVISION"
  )
  
  # Return only the needed variables
  variable_data$variables[needed_vars]
}

#Testing
variable_info_2024 <- get_variable_info(2024)

names(variable_info_2024)

# Test the function with 2021
variable_info_2021 <- get_variable_info(2021)

names(variable_info_2021)
```

The function works for both the older and newer years, so I can use it to get the variable information for all four years. I will save everything together in the data folder so I can use it again later without having to pull it every time.

```{r}
# Get the variable information for each year
variable_info_2021 <- get_variable_info(2021)
variable_info_2022 <- get_variable_info(2022)
variable_info_2023 <- get_variable_info(2023)
variable_info_2024 <- get_variable_info(2024)

# Store all four years together in one list
variable_info <- list(
  `2021` = variable_info_2021,
  `2022` = variable_info_2022,
  `2023` = variable_info_2023,
  `2024` = variable_info_2024
)

# Save the variable information in the data folder
saveRDS(variable_info, "data/variable_info.rds")
```

### Parsing API Data

Now I can start working with the actual PUMS data. I am going to start with a simple API request to see what the returned data looks like and then create a function to turn it into a tibble.

```{r}
# Test a basic PUMS API call
pums_response <- GET(
  "https://api.census.gov/data/2024/acs/acs1/pums",
  query = list(
    get = "AGEP,PWGTP,SEX",
    `for` = "state:37",
    key = Sys.getenv("CENSUS_API_KEY")
  )
)

# Check that the request was successful
status_code(pums_response)
```

The API returns the variable names in the first row and the actual data in the rows after that. I will make a small helper function to separate those and return the data as a tibble.

```{r}
# Convert a Census API response into a tibble
response_to_tibble <- function(response) {
  
  # Read the JSON returned by the API
  api_data <- content(
    response,
    as = "parsed",
    type = "application/json",
    encoding = "UTF-8",
    simplifyVector = FALSE
  )
  
  # The first row has the column names
  column_names <- unlist(api_data[[1]])
  
  # Everything after the first row is the actual data
  data_rows <- lapply(api_data[-1], unlist)
  data_matrix <- do.call(rbind, data_rows)
  
  # Turn everything into a tibble and add the column names
  census_tibble <- as_tibble(
    data_matrix,
    .name_repair = "minimal"
  )
  
  names(census_tibble) <- column_names
  
  census_tibble
}
```

The helper function gets the data into a tibble, but the variables still need to be converted to the right types. I will handle that in the main function.

JWAP and JWDP are a little different from the other numeric variables because the Census stores them as codes for ranges of time. I will use the middle of each time range as the value, so I made a helper function to calculate the midpoint.

```{r}
# convert a Census time range into its midpoint
get_time_midpoint <- function(time_range) {
  
  # Return NA if there is no time to use
  if (
    length(time_range) == 0 ||
    is.na(time_range) ||
    time_range == ""
  ) {
    return(NA_real_)
  }
  
  # Clean up a.m. and p.m. so the times can be read
  time_range <- str_replace_all(
    time_range,
    c("a\\.m\\." = "AM", "p\\.m\\." = "PM")
  )
  
  # Split the beginning and ending times
  times <- str_split(time_range, " to ")[[1]]
  
  # Read both times using lubridate
  start_time <- parse_date_time(times[1], orders = "I:M p")
  end_time <- parse_date_time(times[2], orders = "I:M p")
  
  # Find the middle of the time range
  start_minutes <- hour(start_time) * 60 + minute(start_time)
  end_minutes <- hour(end_time) * 60 + minute(end_time)
  
  (start_minutes + end_minutes) / 2
}
```

Next, I will make the main function for pulling the PUMS data. I want the function to let me choose the year, variables, and geographic area while also checking that the inputs are valid. It will also take care of converting the variables to the right types.

```{r}
get_pums_data <- function(
  year = 2024,
  numeric_vars = c("AGEP", "PWGTP"),
  categorical_vars = "SEX",
  geography = "State",
  geography_value = NULL
) {
  
  # Make sure the year is one of the four years being used
  valid_years <- c(2021, 2022, 2023, 2024)
  
  if (!year %in% valid_years) {
    stop("Year must be 2021, 2022, 2023, or 2024.")
  }
  
  # Check the numeric variables
  valid_numeric_vars <- c(
    "AGEP", "GASP", "GRPIP", "JWAP",
    "JWDP", "JWMNP", "PWGTP"
  )
  
  if (!all(numeric_vars %in% valid_numeric_vars)) {
    stop("One or more numeric variables are invalid.")
  }
  
  if (!any(numeric_vars != "PWGTP")) {
    stop("At least one numeric variable besides PWGTP must be selected.")
  }
  
  # PWGTP always needs to be included
  numeric_vars <- unique(c(numeric_vars, "PWGTP"))
  
  # Check the categorical variables
  valid_categorical_vars <- c(
    "FER", "HHL", "SCH", "SCHL", "SEX"
  )
  
  if (length(categorical_vars) == 0) {
    stop("At least one categorical variable must be selected.")
  }
  
  if (!all(categorical_vars %in% valid_categorical_vars)) {
    stop("One or more categorical variables are invalid.")
  }
  
  # Make sure only one geography level was chosen
  valid_geographies <- c("Region", "Division", "State")
  
  if (length(geography) != 1) {
    stop("Exactly one geography level must be selected.")
  }
  
  if (!geography %in% valid_geographies) {
    stop("Geography must be Region, Division, or State.")
  }
  
  # Load the variable information saved earlier
  variable_info <- readRDS("data/variable_info.rds")
  year_info <- variable_info[[as.character(year)]]
  
  # Set up the selected geography
  if (geography == "State") {
    geography_var <- if (year %in% c(2021, 2022)) "ST" else "STATE"
    geography_values <- year_info[[geography_var]]$values$item
    default_value <- "37"
    
  } else if (geography == "Region") {
    geography_var <- "REGION"
    geography_values <- year_info$REGION$values$item
    default_value <- "3"
    
  } else {
    geography_var <- "DIVISION"
    geography_values <- year_info$DIVISION$values$item
    default_value <- "5"
  }
  
  # Use the default if a specific geography was not given
  if (is.null(geography_value)) {
    geography_value <- default_value
  }
  
  # Make sure the geography value is actually available
  if (
    geography_value != "all" &&
    !geography_value %in% names(geography_values)
  ) {
    stop("The selected geography value is invalid.")
  }
  
  # Put together the variables we want from the API
  requested_vars <- unique(c(
    numeric_vars,
    categorical_vars,
    geography_var
  ))
  
  # Build the API URL for the selected year
  api_url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums"
  )
  
  # Census uses * when we want every value for a geography
  api_geography_value <- if (
    geography_value == "all"
  ) {
    "*"
  } else {
    geography_value
  }
  
  geography_query <- paste0(
    tolower(geography),
    ":",
    api_geography_value
  )
  
  # Pull the data from the Census API
  response <- GET(
    api_url,
    query = list(
      get = paste(requested_vars, collapse = ","),
      `for` = geography_query,
      key = Sys.getenv("CENSUS_API_KEY")
    )
  )
  
  # Turn the API response into a tibble
  census_data <- response_to_tibble(response)
  
  # Convert the regular numeric variables
  regular_numeric_vars <- setdiff(
    numeric_vars,
    c("JWAP", "JWDP")
  )
  
  census_data <- census_data |>
    mutate(
      across(
        all_of(regular_numeric_vars),
        as.numeric
      )
    )
  
  # See if either of the time variables was selected
  time_vars <- intersect(
    numeric_vars,
    c("JWAP", "JWDP")
  )
  
  # Convert the time codes to the middle of each time range
  for (var in time_vars) {
    
    time_values <- year_info[[var]]$values$item
    
    time_ranges <- sapply(
      census_data[[var]],
      function(code) {
        
        # A code of 0 means there is no time to use
        if (is.na(code) || code == "" || code == "0") {
          return(NA_character_)
        }
        
        # Add leading zeros so the codes match the saved information
        code <- str_pad(
          code,
          width = 3,
          side = "left",
          pad = "0"
        )
        
        # If there still is not a match, treat it as missing
        if (!code %in% names(time_values)) {
          return(NA_character_)
        }
        
        unlist(time_values[[code]])
      }
    )
    
    census_data[[var]] <- sapply(
      time_ranges,
      get_time_midpoint
    )
  }
  
  # Replace the category codes with their actual labels
  for (var in categorical_vars) {
    
    category_values <- year_info[[var]]$values$item
    
    census_data[[var]] <- factor(
      census_data[[var]],
      levels = names(category_values),
      labels = unname(unlist(category_values))
    )
  }
  
  # The API adds the geography again, so remove that extra column
  api_geography_var <- tolower(geography)
  
  if (
    api_geography_var %in% names(census_data) &&
    api_geography_var != geography_var
  ) {
    census_data <- census_data |>
      select(-all_of(api_geography_var))
  }
  
  # Give the finished tibble the census class
  class(census_data) <- c(
    "census",
    class(census_data)
  )
  
  census_data
}
```
### Combining Years

The function above only pulls one year at a time, but I also want to be able to look at multiple years together. I will make another function that uses the single-year function for each year I choose and then combines the results. I will also add a year column so I can tell which year each observation came from.

```{r}
# Pull PUMS data for more than one year
get_multiple_years <- function(
  years = "all",
  numeric_vars = c("AGEP", "PWGTP"),
  categorical_vars = "SEX",
  geography = "State",
  geography_value = NULL
) {
  
  # These are the years available for this project
  valid_years <- c(2021, 2022, 2023, 2024)
  
  # Use all four years if "all" is selected
  if (length(years) == 1 && years == "all") {
    years <- valid_years
  }
  
  # Make sure all of the selected years are available
  if (!all(years %in% valid_years)) {
    stop("Years must be 2021, 2022, 2023, 2024, or all.")
  }
  # Pull the data for each selected year
year_data <- lapply(
  years,
  function(current_year) {
    
    data <- get_pums_data(
      year = current_year,
      numeric_vars = numeric_vars,
      categorical_vars = categorical_vars,
      geography = geography,
      geography_value = geography_value
    )
    
    # Keep track of which year each row came from
    data$year <- current_year
    
    data
  }
)

# Combine all of the years into one tibble
combined_data <- bind_rows(year_data)

# Treat year as a category instead of a numeric variable
combined_data$year <- factor(combined_data$year)

# Keep the census class after combining the data
class(combined_data) <- c(
  "census",
  setdiff(class(combined_data), "census")
)

combined_data
}
```

## Part 2: Generic Functions

### Summary Function

Now that the data can be pulled and combined, I want an easier way to summarize it. I will create a summary function specifically for census objects. For numeric variables, I will use the person weights to calculate weighted means and standard deviations. For categorical variables, I will use the weights to get counts for each category.

```{r}
# Summarize a census data set
summary.census <- function(
  object,
  numeric_vars = NULL,
  categorical_vars = NULL,
  ...
) {
  
  # Find the variables we can summarize
  available_numeric <- names(object)[
    sapply(object, is.numeric)
  ]
  
  available_categorical <- names(object)[
    sapply(object, is.factor)
  ]
  
  # PWGTP is the weight, so we do not summarize it itself
  available_numeric <- setdiff(
    available_numeric,
    "PWGTP"
  )
  # Use all available variables if none are selected
if (is.null(numeric_vars)) {
  numeric_vars <- available_numeric
}

if (is.null(categorical_vars)) {
  categorical_vars <- available_categorical
}

# Make sure the selected numeric variables are available
if (!all(numeric_vars %in% available_numeric)) {
  stop("One or more numeric variables are not available.")
}

# Make sure the selected categorical variables are available
if (!all(categorical_vars %in% available_categorical)) {
  stop("One or more categorical variables are not available.")
}
  # Make a list to hold all of the summaries
summary_results <- list()

# Calculate the weighted summaries for numeric variables
for (var in numeric_vars) {
  
  values <- object[[var]]
  weights <- object$PWGTP
  
  # Only use rows that have a value for this variable
  keep <- !is.na(values)
  values <- values[keep]
  weights <- weights[keep]
  
  # Calculate the weighted mean
  weighted_mean <- sum(values * weights) / sum(weights)
  
  # Calculate the weighted standard deviation
  weighted_sd <- sqrt(
    sum(values^2 * weights) / sum(weights) -
      weighted_mean^2
  )
  
  summary_results[[var]] <- c(
    mean = weighted_mean,
    sd = weighted_sd
  )
}
# Get the weighted counts for categorical variables
for (var in categorical_vars) {
  
  values <- object[[var]]
  weights <- object$PWGTP
  
  # Add up the weights for each category
  weighted_counts <- tapply(
    weights,
    values,
    sum,
    na.rm = TRUE
  )
  
  summary_results[[var]] <- weighted_counts
}

# Return all of the summaries
summary_results
}
```

### Plot Function

I also want a quick way to compare a numeric variable across different categories. I will create a plot function for census objects that makes a weighted boxplot using the person weights.

```{r}
# Make a weighted boxplot for census data
plot.census <- function(
  x,
  categorical_var,
  numeric_var,
  ...
) {
  
  # Make sure only one variable of each type was selected
  if (length(categorical_var) != 1 || length(numeric_var) != 1) {
    stop("Choose one categorical variable and one numeric variable.")
  }
  
  # Make sure the categorical variable is in the data
  if (
    !categorical_var %in% names(x) ||
    !is.factor(x[[categorical_var]])
  ) {
    stop("The categorical variable must be a factor in the data.")
  }
  
  # Make sure the numeric variable is in the data
  if (
    !numeric_var %in% names(x) ||
    !is.numeric(x[[numeric_var]]) ||
    numeric_var == "PWGTP"
  ) {
    stop("The numeric variable must be an available numeric variable.")
  }
  # Make the weighted boxplot
ggplot(
  x,
  aes(
    x = get(categorical_var),
    y = get(numeric_var),
    weight = PWGTP
  )
) +
  geom_boxplot() +
  labs(
    x = categorical_var,
    y = numeric_var
  )
}
```


