---
title: "ST 558 Project 1"
author: "Grace Haley"
format: html
editor: visual
---
## Introduction

For this project, I will be working with data from the Public Use Microdata 
Sample (PUMS) Census API. I will create functions to pull and process the 
data, summarize different variables, and create plots. The functions will 
also allow different years and geographic areas to be selected.

## Part 1: Data Processing

### Parsing Variable Values

Before pulling data from the PUMS API, I first need to get information about 
the variables available for each year. The Census API provides JSON files 
that contain the variable descriptions and their possible values. I will 
use these to create lists that can be used later when processing the data.

```{r}
library(httr)
library(jsonlite)
library(tidyverse)
library(lubridate)
```

I will start with the 2024 data to look at how the variable information is 
returned before creating a function that can do this for each year.

```{r}
url_2024 <- "https://api.census.gov/data/2024/acs/acs1/pums/variables.json"

response_2024 <- GET(url_2024)

response_2024

# Parse the JSON response into an R list
variables_2024 <- content(response_2024, as = "parsed")

# Look at the names of the first few variables
head(names(variables_2024$variables))

# Look at the information stored for the SEX variable
variables_2024$variables$SEX

# Pull out the possible values for SEX
variables_2024$variables$SEX$values$item

# Variables needed for 2024
needed_vars <- c(
  "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
  "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
  "STATE", "REGION", "DIVISION"
)

# Keep only the variables needed for the project
selected_2024 <- variables_2024$variables[needed_vars]

names(selected_2024)
```
Now that I know how the variable information is stored, I can put these 
steps into a function. The function will take a year, pull the variable 
information for that year, and keep only the variables needed for this 
project. The state variable is named ST in 2021 and 2022 and STATE in 
2023 and 2024, so I also have to account for that difference.

```{r}
# Pull and parse the variable information for a given year
get_variable_info <- function(year) {
  
  # Create the URL for the selected year
  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums/variables.json"
  )
  
  # Pull and parse the JSON data
  response <- GET(url)
  variable_data <- content(response, as = "parsed")
  
  # The state variable has a different name in 2021 and 2022
  state_var <- if (year %in% c(2021, 2022)) "ST" else "STATE"
  
  # Variables needed for the project
  needed_vars <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
    "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
    state_var, "REGION", "DIVISION"
  )
  
  # Return only the needed variables
  variable_data$variables[needed_vars]
}

#Testing
variable_info_2024 <- get_variable_info(2024)

names(variable_info_2024)

# Test the function with 2021
variable_info_2021 <- get_variable_info(2021)

names(variable_info_2021)
```

The function worked for both 2021 and 2024, including the different state 
variable names. I can now use the function to get the variable information 
for all four years and save the lists together so they can be used later.

```{r}
# Get the variable information for each year
variable_info_2021 <- get_variable_info(2021)
variable_info_2022 <- get_variable_info(2022)
variable_info_2023 <- get_variable_info(2023)
variable_info_2024 <- get_variable_info(2024)

# Store all four years together in one list
variable_info <- list(
  `2021` = variable_info_2021,
  `2022` = variable_info_2022,
  `2023` = variable_info_2023,
  `2024` = variable_info_2024
)

# Save the variable information in the data folder
saveRDS(variable_info, "data/variable_info.rds")
```